---
# This file generates AGENTS.md in the package root.
# It is not part of the package build or the pkgdown site.
# To regenerate it, run from the package root:
#   rmarkdown::render("vignettes/_AGENTS.Rmd", output_file = file.path("AGENTS.md"), output_dir = ".")
output:
  github_document:
    html_preview: false
params:
  url: "https://raw.githubusercontent.com/insightsengineering/teal/refs/heads/main/SHARED_AGENTS.md"
---
# teal.data R Package Development Guide

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Package Overview

`teal.data` is part of the `teal` framework and provides an API to manage datasets keys in a `teal` app.
The keys are column names of datasets that allow merging the different datasets.

It provides 4 main features to the framework:

- `join_keys` to show and modify relationships between datasets.
- `parents()` to modify parent-child relationships.
- `teal_data` object that extends on `qenv` API by adding API to relate datasets.
- See `@vignettes/teal-data.Rmd` as reference material.

## Development Context

The main feature of this package is to provide `teal` apps with functionality to relate between datasets.

### Relationships with other packages

Direct dependencies:

- `teal.code`: `teal_data` extends a `qenv` object from `teal.code`,
  where the code execution and reproducibility features are implemented. 
  Any issue with code execution and reproducibility should be addressed in this package.
  - Any issue with code execution and reproducibility should be addressed in this package

Usage in other framework packages:

- `teal`: uses `teal_data` to manage the data and relationships. 
  - It is passed to the modules which convert it to `teal_reporter` for the report

### Extending teal.data

`teal.data` should allow for its functionality to be extended by users according to their specific needs.
This can be achieved by adding new methods or overwriting default ones


```{r teal_content, echo=FALSE, results='asis'}
# Append common shared instructions from teal framework
common_agents_url <- params$url
common_agents_path <- tempfile()
download.file(url = common_agents_url, destfile = common_agents_path, quiet = TRUE)
common_agents_shared <- readLines(common_agents_path)
# remove a heading line only intented to aovid a markdown lint error
line_to_avoid_lintr <- grep("^# ", common_agents_shared)
common_agents_shared <- common_agents_shared[-1 * line_to_avoid_lintr]
cat(common_claude_shared, sep = "\n")
```
