---
# This file generates AGENTS.md in the package root.
# It is not part of the package build or the pkgdown site.
# To regenerate it, run from the package root:
#   rmarkdown::render("vignettes/_AGENTS.Rmd", output_file = file.path("AGENTS.md"))
output:
  github_document:
    html_preview: false
params:
  url: "https://raw.githubusercontent.com/insightsengineering/teal/refs/heads/add_common_claude%40main/SHARED_AGENTS.md"
---
# teal.data R Package Development Guide

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Package Overview

`teal.data` is part of the `teal` framework and provides an API to manage datasets keys in a `teal` app.
The keys are column names of datasets that allow merging the different datasets.

It provides 4 main features to the framework:

- `join_keys` to show and modify relationships between datasets.
- `parents()` to modify parent-child relationships.
- `teal_data` object that extends on `qenv` API by adding API to relate datasets.
- See `vignettes/teal-data.Rmd` as reference material.

## Development Context

The main feature of this package is to provide `teal` apps with functionality to relate between datasets.

### Relationships with other packages

Direct dependencies:

- `teal.code`: `teal_data` extends a `qenv` object from `teal.code`, where the code execution and reproducibility features are implemented.
  - Any issue with code execution and reproducibility should be addressed in this package

Usage in other framework packages:

- `teal`: uses the API in `teal.reporter` to maintain an instance of the reporter and uses the exported shiny modules for the interface
  - Converts the `data` argument in `teal::init()` function to the `teal_report` object that is used in the modules
  - `teal_report` depends on `teal_data`
- teal module: The `data` argument passed on to modules uses the `teal_report` data type.
  - Automatically tracks the code execution and output objects
  - It is used in custom modules as well as R packages on CRAN: `teal.modules.clinical` and `teal.modules.general`

### Extending teal.data

`teal.data` should allow for its functionality to be extended by users according to their specific needs.
This can be achieved by adding new methods or overwriting default ones


```{r teal_content, echo=FALSE, results='asis'}
# Append common shared instructions from teal framework
common_claude_url <- params$url
common_claude_path <- tempfile()
download.file(url = common_claude_url, destfile = common_claude_path, quiet = TRUE)
common_claude_shared <- readLines(common_claude_path)
cat(common_claude_shared, sep = "\n")
```
