---
title: "Delayed Data Objects"
author: "Dawid Kałędkowski"
date: "15.05.2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Delayed Data Objects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

## Basic understanding

Delayed data objects are `R` objects that contain instructions on how to acquire data. In practice, you will pass these `DDL` objects with its instructions into a `teal` application so that you can launch a `teal` app first and then pull the data afterwards.

### Key concepts

* A **`TealDatasetConnector`** is an object used to pull a single delayed data set into a `teal` app. Connectors to pull data from proprietary data storage are not included in this package.

```{r}
library(teal.data)
library(scda)

# specialized function to create delayed data using scda (could also set keys when using cdisc_dataset_connector)
adlb <- scda_dataset_connector(
  "ADLB", 
  "adlb", 
  keys = get_cdisc_keys("ADLB")
)

# generalized function to create delayed data from code - see package help for other connectors
x <- code_dataset_connector(
  dataname = "ADSL",
  keys = get_cdisc_keys("ADSL"),
  code = "library(scda)\nADSL <- synthetic_cdisc_data(\"latest\")$adsl"
)
```

* A **`TealDataConnector`** is an object used to pull a set of delayed data sets into a `teal` app which all share a common connection (see [Delayed Data Loading](delayed-data-loading.html) for the definition of a connection object).

```{r}
# using scda
adsl <- scda_cdisc_dataset_connector(dataname = "ADSL", scda_dataname = "adsl")
adae <- scda_cdisc_dataset_connector(dataname = "ADAE", scda_dataname = "adae")
adsl_adae <- relational_data_connector(
  connection = data_connection(),
  connectors = list(adsl, adae)
)
```

* The `cdisc_data` function takes a set of **`TealDataConnector`**, **`TealDatasetConnector`** and / or `cdisc_datasets` (non-delayed datasets) to create the **`TealData`** object which is used to create `teal` applications.

```{r}
# create a TealDatasetConnector for ADVS
advs <- scda_cdisc_dataset_connector(dataname = "ADVS", scda_dataname = "advs")
# use cdisc_data() to create a `DDL` object
delayed_data <- cdisc_data(adsl_adae, advs)
```

### Constructors

Below is a list of all of the constructors available in `teal.data` to create `TealDataset` and delayed `TealDatasetConnector` objects:

|                            |                                    Description                |           Base Constructor                | Constructor Wrappers                                                                                                                                                                                                                           |
|:--------------------------:|:-------------------------------------------------------------:|:-----------------------------------------:|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `TealDataset`                | Dataframe with name (and optionally keys)                     | `dataset`, `dataset_file`                      | `dataset`, `cdisc_dataset`                                                                                                                                                                                                                         |
| `TealDatasetConnector`       | Delayed Dataset                                               | `dataset_connector`, `dataset_connector_file`  | (see note 1 below) `rds_dataset_connector`, `script_dataset_connector`, `code_dataset_connector`,  `csv_dataset_connector`, `fun_dataset_connector`, `python_dataset_connector`, `scda_dataset_connector` |
| `TealDataConnector`          | Group of `TealDatasetConnector`                                 |                   |                                                                                                                                                                                                                                                |
| `TealData`                   | Group of `TealDatasetConnector`, `TealDataConnector`, `TealDataset` | `teal_data`, `teal_data_file`                  | (see note 2 below) `cdisc_data`, `cdisc_data_file`                                                                                                                                                                                                  |

Notes:

1. All `xyz_dataset_connector` functions have an equivalent `xyz_cdisc_dataset_connector` function (for example `rds_cdisc_dataset_connector`) which specifies additional dataset metadata.
2. `cdisc_data` is the standard function used to create a data object to be used within teal apps for standard `CDISC` study data. The more general `teal_data` function can be used to allow arbitrary relational data to be used within teal apps.

### Dataset dependencies

The datasets passed into `teal_data` and `cdisc_data` are pulled in the order they are inputted. So if datasets depend on other datasets being available they should be placed later in the argument list:

```{r}
adsl <- scda_cdisc_dataset_connector(dataname = "ADSL", "adsl")
adsl_2 <- code_cdisc_dataset_connector("ADSL_2",
  code = "head(ADSL, 5)",
  keys = get_cdisc_keys("ADSL"), ADSL = adsl
)

# launch method will be able to load the data as adsl will be pulled first
cdisc_data(adsl, adsl_2)

# launch method will not be able to load the data as adae is pulled first but it depends on adsl
cdisc_data(adsl_2, adsl)
```

## Suggested development workflow

The following workflow facilitates building `teal` apps with `DDL` by minimizing debugging overhead.

1. Run a teal app configured without delayed data to verify that the app starts as expected.
1. Replace the `cdisc_dataset` functions with the appropriate `TealDatasetConnector` objects.
1. Add preprocessing code (see [Delayed Data Loading](delayed-data-loading.html) for preprocessing documentation) and verify once again by running the newly created object's `$launch` method.
1. First, include just a single module to verify that the teal app starts and that it loads all the expected data.
1. Then, add the rest of the modules one by one, replacing all dataset calls with strings and iteratively verify that the app functions as expected.
