---
title: "Join Keys"
author: "NEST CoreDev"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Join Keys}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

The `teal.data` package provides a way to define primary keys for a dataset and establish relationships with other datasets.

Each dataset can be characterized by:

- Columns constituting the primary key
- Merge keys, analogous to `SQL` foreign keys

Typically, an application developer specifies these keys manually. However, for datasets following the `ADaM` standard, `teal.data` can automatically assign keys using the `default_cdisc_join_keys` object.
Refer to the section ["Joining Keys with `ADaM` Datasets"](#join-keys-in-adam-datasets) for details on using this object to select specific datasets.

##### Uses of _join_keys_ class in _teal_ applications

The primary function of `join_keys` in `teal` applications is to facilitate the seamless [merging of datasets](https://insightsengineering.github.io/teal.transform/latest-tag/articles/data-merge.html) using `teal.transform`.

Additionally, it plays a role on the data filtering using the _[Filter Panel](https://insightsengineering.github.io/teal/main/articles/filter-panel.html)_ in a `teal` application.
The filters applied to a _(parent)_ dataset are also applied to their children.

## Anatomy of Join Keys

The `join_keys` object contains information about the foreign/primary keys of multiple datasets.
Each key is represented by a pair of datasets _(by name reference)_ and a named character vector that encodes the match column name between the two datasets.
In addition, a foreign key also contains a _parent-child_ attribute that is used in the "Filter Panel" as we mentioned above.

A new `join_keys` can be created as empty or by defining an initial set of primary and foreign keys.
That initial object can be extended by adding/modifying/removing keys and by establishing parent-child relationships between datasets.

The `join_keys` function is used both as a constructor and as a getter.
As the getter it is used to retrieve the `join_keys` that is contained in other objects, such as `teal_data`.

As the constructor it is used to specify keys `join_keys(...)` as a collection of multiple `join_key` entries:

- `join_key(dataset_1, dataset_2, key)`: specifies the relationship between two datasets:
  - `dataset_1`, `dataset_2`: names of the datasets (if `dataset_2` is the same as `dataset_1`, it creates a primary key)
  - `key` _(optional)_: named vector of column names

Note that join keys are assumed to be symmetric, i.e., `join_key("ds1", "ds2", c("ds1_col" = "ds2_col"))` establishes a relationship from "x" to "y" and vice versa.

```{r, results="hide", message=FALSE, tidy=FALSE}
library(teal.data)
jk <- join_keys(
  join_key("ds1", keys = "col_1"),                    # ds1: [col_1]
  join_key("ds2", keys = c("col_1", "col_2")),        # ds2: [col_1, col_2]
  join_key("ds3", keys = c("col_1", "col_3")),        # ds3: [col_1, col_3]
  join_key("ds1", "ds2", keys = "col_1"),             # ds1 <--> ds2
  join_key("ds1", "ds3", keys = "col_1"),             # ds1 <--> ds3
  join_key("ds4", "ds5", keys = c("col_4" = "col_5")) # ds4 <--> ds5
)

# The parent of ds2 and ds3 is ds1
# converts relationship to child-parent
#  ds1 <--> ds2 becomes ds1 <-- ds2
#  ds1 <--> ds3 becomes ds1 <-- ds3
parents(jk) <- list(ds2 = "ds1", ds3 = "ds1")

jk
```

| Output of `print(jk)`              | Output annotation                    |
| ---------------------------------- |:----------------------------------------:|
| `## A join_keys object containing foreign keys between 3 datasets:` |  **Title** |
| `## ds1: [col_1]`                |  **Primary keys**<br />_\<dataset name\>_: [_\<primary_keys\>_]                                          | 
| `##   <-- ds2: [col_1]`          | **Foreign keys**<br /> _(arrow `<--` denotes `ds1` is the parent of `ds2`)_                    | 
| `##   <-- ds3: [col_1]`          |                                                                               | 
| `## ds2: [col_1, col_2]`           |                                                                               | 
| `##   --> ds1: [col_1]`          | arrow `-->` denotes `ds2` is a child of `ds1`                                                                              | 
| `##   --* (implicit via parent with): ds3` |  **Implicit relationship between `ds2` & `ds3`**<br />_(given that they share common keys with same parent)_    |
| `## ds3: [col_1, col_3]`           |                                                                               | 
| `##   --> ds1: [col_1]`          |                                                                               | 
| `##   --* (implicit via parent with): ds2` |                                                                               | 
| `## ds4: [no primary keys]`    |                                                                               | 
| `##   <-> ds5: [col_5]`          |  **Foreign keys** <br />_(arrow `<->` denotes no parent definition between datasets)_      | 
| `## ds5: [no primary keys]`    |                                                                               | 
| `##   <-> ds4: [col_4] `         |                                                                               | 

## Accessing and Modifying keys

The _subset_ operator with 2 indices is used to retrieve the primary/foreign keys in a `join_keys`.
Both indices must be a string.

```{r}
# Using the jk object defined on "Anatomy of Join Keys"
jk

# Getting primary key of "ds1"
jk["ds1", "ds1"]

# Getting foreign keys between "ds4" and "ds5"
jk["ds4", "ds5"]
```

Note that there is a symmetry between `ds4` and `ds5` relationship:

```{r}
jk["ds5", "ds4"]
jk["ds5", "ds4"]
```


When only 1 argument is used this operator will return a `join_keys` object that is filtered accordingly.

```{r}
# Using the jk object defined on "Anatomy of Join Keys"
jk

# Getting primary key of "ds1"
jk["ds1"]
```


Modifying or adding a key uses the same notation with the assignment operator `<-`.
A symmetric relationship will be created automatically.
Assigning `NULL` value will delete the relationship.

```{r}
# Adding a new ds5 <-> ds1 key
jk["ds5", "ds1"] <- "a_column"

# Removing an existing key
jk["ds4", "ds5"] <- NULL
```

## Join Keys Relationships

There are 2 types of relationships encoded with `join_keys` described in the following sections. The _primary_ and _foreign_ keys are created explicitly using the `join_key` function.

Additionally, the `join_keys` object detects implicit relationships when two datasets share foreign keys to a parent dataset, but not between themselves. These implicit relationships are available just like another foreign key and can be used to merge datasets, despite not being defined by the user.

### Primary Key with `teal_data`

When using the `teal_data` function, the simplest method to define the join keys  is to use the `join_keys` argument.
We can specify the column(s) of the dataset that _(together)_ uniquely identify rows in the dataset.

```{r, include=FALSE}
# nolint start: commented_code_linter.
```

```{r, message=FALSE}
library(teal.data)

td_pk <- within(
  teal_data(),
  ds1 <- dplyr::mutate(iris, id = factor(dplyr::row_number()))
)
datanames(td_pk) <- "ds1"

join_keys(td_pk) <- join_keys(join_key("ds1", keys = "id"))

join_keys(td_pk)
```

We can extend the previous example and define primary keys for multiple datasets:

```{r, message=FALSE}
library(teal.data)

td_pk <- within(
  td_pk,
  {
    ds2 <- data.frame(W = 10:1, V = 5:14, M = rep(1:5, 2))
    ds3 <- data.frame(V = 5:14, N = 4)
  }
)
datanames(td_pk) <- c(datanames(td_pk), "ds2", "ds3")

join_keys(td_pk)["ds2", "ds2"] <- c("V", "W")
join_keys(td_pk)["ds3", "ds3"] <- c("V", "W")

join_keys(td_pk)
```

```{r, include=FALSE}
# nolint end: commented_code_linter.
```

### Foreign Keys with `teal_data`

When passing multiple datasets to the `teal_data` function, dataset relationships are set using `join_keys` and `join_key` and these are used to merge datasets together within `teal` apps.
For users familiar with `SQL` database schema, these relationships are symmetric and not as strict as `SQL` foreign key relationships as `teal` does not validate whether the values inserted into foreign key columns are present in the parent table.

For example:

```{r}
library(teal.data)

td_fk <- within(
  teal_data(),
  {
    ds1 <- data.frame(X = 1:10, Y = 21:30, Z = 1:10)
    ds2 <- data.frame(W = 10:1, V = 5:14, M = rep(1:5, 2))
    ds3 <- data.frame(V = 5:14, N = 4)
  }
)
datanames(td_fk) <- c("ds1", "ds2", "ds3")

join_keys(td_fk) <- join_keys(
  # Primary keys
  join_key("ds1", keys = c("X")),
  join_key("ds2", keys = c("V", "W")),
  join_key("ds3", keys = c("V")),
  # Foreign keys
  join_key("ds2", "ds1", c("W" = "X")),
  join_key("ds3", "ds2", c("V" = "V"))
)

# The parent of ds2 and ds3 is ds1
parents(td_fk) <- list(ds2 = "ds1", ds3 = "ds2")

join_keys(td_fk)
```

### Implicit Relationships

Two datasets that share common _foreign_ keys to the same _parent_ dataset have an implicit relationship between them that is modeled and accessible in `join_keys`.

This is a special inferred relationship from existing `join_keys` that does not need to be explicitly defined and can be seamlessly accessible just as any other foreign key.
As any other foreign key it can be overwritten.

These implicit relationships can be used to merge 2 datasets together, just as if they were defined manually.

```{r}
library(teal.data)

td <- within(
  teal_data(),
  {
    ds1 <- data.frame(X = 1:10, Y = 21:30, Z = 1:10)
    ds2 <- data.frame(W = 10:1, V = 5:14, M = rep(1:5, 2))
    ds3 <- data.frame(V = 5:14, N = 4)
    ds4 <- data.frame(V = 5:14, R = rnorm(10))
  }
)
datanames(td) <- c("ds1", "ds2", "ds3", "ds4")

join_keys(td) <- join_keys(
  # Primary keys
  join_key("ds1", keys = c("X")),
  join_key("ds2", keys = c("V", "W")),
  join_key("ds3", keys = c("V")),
  join_key("ds4", keys = c("V")),
  # Foreign keys
  join_key("ds2", "ds1", c("W" = "X")),
  join_key("ds3", "ds2", c("V" = "V")),
  join_key("ds4", "ds1", c("V" = "X"))
)

# The parent of ds2 and ds3 is ds1
parents(td) <- list(ds2 = "ds1", ds3 = "ds2", ds4 = "ds1")

join_keys(td)

join_keys(td)["ds2", "ds4"]
```

Note that the definition above contains no `join_key` for `"ds2" <-> "ds4"`

## _Join Keys_ in `ADaM` Datasets

`teal.data` provides a set of default join keys (primary and foreign) for datasets named according to the `ADaM` standard. They are stored in `default_cdisc_join_keys`.

```{r}
names(default_cdisc_join_keys) |> sort()
```

When not all default keys are required users can select a smaller set of datasets by subsetting (for example: `default_cdisc_join_keys[c("ADSL", "ADTTE")]`).
Please note, that for every element that is selected, its parent will also be returned (if it has one) as well as any pair-wise relationships in the resulting selection.

```{r}
default_cdisc_join_keys

default_cdisc_join_keys["ADSL"]

default_cdisc_join_keys["ADTTE"]

default_cdisc_join_keys[c("ADSL", "ADTTE", "ADRS")]
```
