---
title: "Join Keys"
author: "NEST CoreDev"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Join Keys}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview
`teal.data` gives the possibility to define individual keys per dataset and to define the relations to other datasets.

Each table can have a set of keys that inform its structure and relation to other tables by specifying:

- columns consisting the primary key
- merge keys, a concept similar to `SQL`'s foreign key.

Usually, an application developer needs to specify the keys manually, but in case of datasets named according
to the `ADaM` standard, `teal` can assign the keys on its own. 
Refer to `vignette("including-adam-data-in-teal", package = "teal")` for more information.

### Primary key

Using the `keys` argument to the `dataset` function (or for `DDL` a `dataset_connector`), we can specify the column(s) 
of the dataset that (together) uniquely identify rows in the dataset. 

```{r, message=FALSE}
library(teal.data)
library(dplyr)

iris_with_keys <- mutate(iris, id = factor(row_number()))

ds <- dataset(
  "IRIS",
  iris_with_keys,
  code = "IRIS <- iris %>% mutate(id = factor(row_number()))",
  keys = c("id")
)

ds$get_keys()
```

### Merge keys

When passing multiple datasets to the `cdisc_data` function, dataset relationship are set using
`join_keys` and `join_key` and these are used to merge datasets together
within `teal` apps. For users familiar with `SQL` database schema, these relationships are symmetric and not as strict
as `SQL` foreign key relationships as `teal` does not validate whether the values inserted into foreign key
columns are present in the parent table.

For example:

```{r, eval=FALSE}
library(teal.data)

data_1 <- data.frame(X = factor(1:10), Y = 21:30, Z = 1:10)
data_2 <- data.frame(W = factor(10:1), V = factor(5:14), M = rep(1:5, 2))
data_3 <- data.frame(V = factor(5:14), T = 4)

data <- teal_data(
  dataset("D1", data_1, code = "D1 <- data.frame(X = factor(1:10), Y = 21:30, Z = 1:10)"),
  dataset("D2", data_2, code = "D2 <- data.frame(W = factor(10:1), V = factor(5:14), M = rep(1:5, 2))"),
  dataset("D3", data_3, code = "D3 <- data.frame(V = factor(5:14), T = 4)"),
  join_keys = join_keys(
    join_key("D1", "D2", c("X" = "W")),
    join_key("D2", "D3", c("V" = "V"))
  )
)
```

The `join_keys` function is used to specify keys:

- `join_keys` is a collection of multiple `join_key` entries
- `join_key` specifies the relation between two datasets:
  - `dataset_1`, `dataset_2` - name of two datasets
  - `key` - (optionally) named vector of column names

Note that it is assumed that join keys are symmetric, i.e. `join_key("x", "y", "x_col" = "y_col")` will enable merge
from "x" to "y" and vice versa.
