---
title: "Pre-processing data"
author: "NEST CoreDev"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pre-processing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

"Data" pre-processing refers to the code which contains:

1. Data import calls
2. Data modification

Including the pre-processing code is an important step that is handled by `teal.data` functions.

For example, for the following example only `ADSL <- readRDS("<your data path>/adsl.rds")` is considered as pre-processing code:

```{r, message=FALSE}
library(teal.data)
saveRDS(example_cdisc_data("ADSL"), "adsl.rds")

## preprocessing>
adsl <- readRDS("adsl.rds")
## <preprocessing

data <- cdisc_data(adsl = adsl)
teal.code::get_code(data)
```

When executing the example above, the `get_code` function will return an empty string reflecting that _"Pre-processing is empty"_. In order to show the pre-processing code correctly the `code` argument of the `cdisc_data` function needs to be specified. For the example above this would be:

```{r, message=FALSE}
library(teal.data)
library(magrittr)
saveRDS(example_cdisc_data("ADSL"), "adsl.rds")

## preprocessing>
adsl <- readRDS("adsl.rds")
## <preprocessing
unlink("adsl.rds")


data <- cdisc_data(
  adsl = adsl,
  code = quote(ADSL <- readRDS("adsl.rds"))
)
teal.code::get_code(data)
```

The code used to get the `ADSL` dataset is returned as expected. This can be used as input to the `data` argument of `teal::init` to ensure reproducibility in `teal` apps.
