---
title: "Reproducibility check"
author: "NEST CoreDev"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducibility check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Reproducibility check

## Mechanism

`teal.data` classes that accumulate raw data accept a boolean `check` argument, e.g. `teal_data`. This boolean
flag indicates whether `teal` performs validation of the data passed to the `TealData` object during launch of a
`teal` application.

The validation compares the raw data passed to the class with the dataset returned by the code stored in the class.
E.g.:

```{r, message=FALSE}
library(teal.data)
data_verified <- within(teal_data(), iris$id <- seq_len(nrow(iris)))

data_unverified <- teal_data(iris = iris)
within(data_unverified, iris$id <- seq_len(nrow(iris)))
# verify(data)
```

The `check()` method called during launch can be invoked directly and returns `TRUE` if the raw data passed
to `teal_data` matches the return value of the code stored in the class. It returns an error if the comparison
fails and `NULL` if the `check = FALSE` was passed to `teal_data`.

```{r}
data <- teal_data(iris = iris, code = "iris <- mtcars")

# returns NULL even though the code does not match the raw data
# data$check()
```

Find out more in the `teal_data()` documentation.

## Final considerations
- `<x>_connector` objects do not accept raw data as input and therefore are not checked even if `check = TRUE` is used.
