---
title: "Introduction to teal.data"
author: "NEST CoreDev"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to teal.data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The `teal.data` package enables `teal` application developers to convert their data into a format which can be used
inside `teal` applications.

## Quick Start

For creating `ADaM` clinical trial data please use the `cdisc_data` and `cdisc_dataset` functions:

```{r, message=FALSE}
library(teal.data)

# create some dummy data
adsl <- teal.data::example_cdisc_data("ADSL")
adtte <- teal.data::example_cdisc_data("ADTTE")

my_data <- cdisc_data(
  cdisc_dataset("ADSL", adsl),
  cdisc_dataset("ADTTE", adtte)
)
```

For more general data use the `teal_data` and `dataset` functions:

```{r}
my_general_data <- teal_data(
  dataset("iris", iris),
  dataset("mtcars", mtcars)
)
```


## Additional functionality

The `teal.data` package provides many features to help specify your data:

- [Specifying `ADaM` data](https://insightsengineering.github.io/teal/latest-tag/articles/including-adam-data-in-teal.html)
- [Specifying `MultiAssayExperiment` data](https://insightsengineering.github.io/teal/latest-tag/articles/including-mae-data-in-teal.html)
- [Specifying general relational data](https://insightsengineering.github.io/teal/latest-tag/articles/including-general-data-in-teal.html)
- [Specifying relationships between your datasets](https://insightsengineering.github.io/teal.data/latest-tag/articles/join-keys.html)
- [Dynamically loading your data (`DDL`)](https://insightsengineering.github.io/teal.data/latest-tag/articles/using-delayed-data-basic.html)
- [Pre-processing data](https://insightsengineering.github.io/teal.data/latest-tag/articles/preprocessing-data.html)
- [Pre-processing `DDL` data ](https://insightsengineering.github.io/teal.data/latest-tag/articles/preprocessing-delayed-data.html)
- [Checking for reproducibility](https://insightsengineering.github.io/teal.data/latest-tag/articles/reproducibility.html)
- [Using python to create your data](https://insightsengineering.github.io/teal.data/latest-tag/articles/teal.data-with-python.html)
