---
title: "Introduction to teal.data"
author: "NEST CoreDev"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to teal.data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The `teal.data` package enables `teal` application developers to convert their data into a format which can be used
inside `teal` applications. `teal_data` class inherits from [`qenv`](https://insightsengineering.github.io/teal.code/latest-tag/articles/qenv.html) and is meant to be used for reproducibility purposes. 

## Quick Start

To create a `teal_data` class object, use the `teal_data` function. `teal_data` has a number of methods which allow 
to set and get relevant information from private class slots.

```{r, results = 'hide'}
library(teal.data)

my_data <- teal_data()
my_data <- within(
  my_data,
  {
    data1 <- data.frame(x = 1:10, y = 11:20)
    data2 <- data.frame(x = 1:10, y = 21:30)
  }
)

# get variables from teal_data environment
my_data[["data1"]]
my_data[["data1"]]

# get reproducible code
get_code(my_data)

# set/get datanames
datanames(my_data) <- c("data1", "data2")
datanames(my_data)

# concise print
print(my_data)
```


## `teal_data` characteristics

`teal_data` class object keeps the following information:
- `data` - an environment containing data. 
- `code` - a string containing code to reproduce the `data` (details in [teal_data reproducibility ](teal-data-reproducibility.html)).
- `datanames` - a character vector with names of datasets.
- `join_keys` - a `join_keys` object defining relationships between datasets (details in [Relational data](relational-data.html)).

### Reproducibility

Primary function of `teal_data` is to provide reproducibility of data. We recommend to initialise empty `teal_data` which marks object as _verified_. Read more in [teal_data Reproducibility](teal-data-reproducibility.html).

```{r}
my_data <- teal_data()
my_data <- within(my_data, IRIS <- iris)
my_data <- within(my_data, IRIS$id <- seq_len(nrow(iris)))
my_data # is verified
```


### Relational data models

`teal_data` class can be extended to support relational data. `join_keys` function can be used to specify relationships between datasets. See more in [join_keys](join-keys.html).

```{r}
my_data <- teal_data(
  parent = data.frame(id = 1:10, y = 11:20),
  child = data.frame(id = 1:20, parent_id = c(1:10, 1:10), y = 21:30)
)

join_keys(my_data) <- join_keys(
  join_key("parent", "parent", key = "id"),
  join_key("child", "child", key = "id"),
  join_key("child", "parent", key = c("parent_id" = "id"))
)

join_keys(my_data)
```
