---
title: "teal.data with Python"
author: "NEST CoreDev"
date: "2022-05-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{teal.data with Python}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

It is possible to use python code (via the `reticulate` package) to create a `teal` dataset. 
Therefore if a python package is required to access/create your data, it can still be used in a `teal` application.

We recommend having a thorough understanding of the `reticulate` package (see [here](https://rstudio.github.io/reticulate/))
before using the python functionality described below. 

## Example

In this section we show a simple example of how python code can be used to create a `teal` dataset.

For this example we require a python environment with `pandas` installed. See [here](https://rstudio.github.io/reticulate/articles/python_packages.html) for further details.

We first load the packages:

```{r check_reticulate, include = FALSE}
reticulate_available <- requireNamespace("reticulate", quietly = TRUE)
```


```{r, message=FALSE, eval = reticulate_available}
library(reticulate) # may need to be installed if not available
# reticulate function to call to install the pandas package if not installed: py_install("pandas")
```


```{r, message=FALSE}
library(teal.data)
```

Next we define the python code which we want to use to create the dataset.
In this example, for demonstration purposes we create a trivial dataset which could easily be done directly in R.
Note we use a variable `num_rows` which needs to be passed to the python code from R. 

```{r, eval = reticulate_available}
python_code <- "import pandas as pd
data = pd.DataFrame({\"id\" : range(r.num_rows), \"val\" : range(r.num_rows)})"
```

Next we create a `python_dataset_connector` object. 

```{r, eval = reticulate_available}
x <- python_dataset_connector(
  dataname = "DATA", # the teal dataset name
  code = python_code, # the code used to generate the dataset
  object = "data", # the object in the python code to be converted to a data.frame for the teal dataset
  keys = "id", # the key for teal dataset object
  vars = list(num_rows = 5L) # any variables passed from R into python (note this could be an R variable)
)
```

Finally we can test the code by pulling the data into R:

```{r eval = FALSE}
x$pull()
print(x)
```

## Further Concerns

### Reproducibility

In order to ensure reproducibility when generating `teal` datasets using python it is necessary to use a reproducible
python environment (for example using `virtualenv` or `conda`) see [here](https://rstudio.github.io/reticulate/articles/versions.html)
for more details.

### Deployment

A python interpreter (or the python environment described above) will need to be available when deploying your app.
We recommend following the example [here](https://github.com/ranikay/shiny-reticulate-app) to see how this can be done.
