---
title: "Examining Delayed Data Objects"
author: "Dawid Kałędkowski"
date: "15.03.2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examining Delayed Data Objects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

## Overview

Once your delayed data object has been created as described in [Delayed Data Objects](using-delayed-data-basic.html), `teal.data` provides a useful set of functions to examine the object outside of a shiny application, i.e. the global environment. Below is an exhaustive list of all such functions:

|                                             | `TealDataset`       | `TealDatasetConnector`       | `TealDataConnector` & `TealData`      |
|---------------------------------------------|-------------------|----------------------------|------------------------------------------|
| Get Reproducible Code (Optionally Deparsed) | `get_code`          | `get_code`                   | `get_code`                          |
| Get data.frame                              | `get_raw_data`      | `get_raw_data`               | `get_raw_data`                      |
| Get Dataset Name                            | `get_dataname`      | `get_dataname`               | `get_dataname`                      |
| Get Single Dataset Object                   | `get_dataset`       | `get_dataset`                | `get_dataset`                       |
| Get All Dataset Objects                     | -                 | -                          | `get_datasets`                      |
| Load Data                                   | -                 | `load_dataset`               | `load_datasets`                     |
| Check if Loaded                             | -                 | `is_pulled`                  | `is_pulled`                         |
| Mutate Single Dataset                       | `mutate_dataset`    | `mutate_dataset`             | `mutate_dataset`                    |
| Mutate All Datasets                         | -                 | -                          | `mutate_data`                       |

The most basic function `get_dataname` returns the name of the dataset or datasets in your delayed data object:

```{r, message=FALSE}
library(teal.data)

adsl_cf <- callable_function(function() example_cdisc_data("ADSL"))
adsl <- cdisc_dataset_connector(
  dataname = "ADSL",
  pull_callable = adsl_cf,
  keys = get_cdisc_keys("ADSL")
)
get_dataname(adsl) # "ADSL"

adae_cf <- callable_function(function() example_cdisc_data("ADAE"))
adae <- cdisc_dataset_connector(
  dataname = "ADAE",
  pull_callable = adae_cf,
  keys = get_cdisc_keys("ADAE")
)
delayed_data <- cdisc_data(adsl, adae)
get_dataname(delayed_data) # "ADSL" "ADAE"
```

The delayed data objects described above all also contain a `launch` method which can be used to test the data
loading screen:

```{r}
if (interactive()) {
  delayed_data$launch()
}
```

There is also a `pull` method to test that the data can be loaded without launching a shiny app. See [Delayed Data Advanced](using-delayed-data-advanced.html).

Alternatively `teal.data` provides a `load_dataset` function for `<...>Dataset<...>` objects which is used to pull the data without launching the delayed loading screen, and a `load_datasets`function for `<...>Data<...>` objects which launches the delayed loading screen used to pull the datasets from the connection.

After loading the data, it can be checked that the data has been successfully pulled using the `is_pulled` function:

```{r}
if (interactive()) {
  load_datasets(delayed_data)
}
is_pulled(delayed_data)
```

#### Aside: Loading page UI

It is possible to set default values of the boxes on the loading page using the `set_ui_input`
method:

```{r}
adae$set_ui_input(function(ns) {
  list(numericInput(inputId = ns("n"), label = "Choose number of records", min = 0, value = 5))
})
```

#### Testing data loading continued

Once the data are loaded, it's also possible to access the individual dataset objects using the `get_dataset` function, or for `<...>Data<...>` objects, retrieve all dataset objects using the `get_datasets` function:

```{r results=FALSE}
lapply(delayed_data$get_items(), function(item) item$pull())

# return a particular dataset by name
get_dataset(delayed_data, dataname = "ADSL")

# or return all datasets
load_datasets(delayed_data)
get_datasets(delayed_data)
```

Note that when a connector is loaded, the result is a dataset object:

```{r}

# "CDISCTealDatasetConnector" "TealDatasetConnector" "R6"
class(adsl)

# "CDISCTealDataset" "TealDataset" "R6"
class(get_dataset(adsl))
```

To view the raw dataframe object, use the `get_raw_data` function:

```{r}
# for a single <...>Dataset<..> object
head(get_raw_data(adsl), 2)

# or for a <...>Data<...> object containing multiple datasets, specify the name of the dataset of interest
raw <- get_raw_data(delayed_data, "ADSL")
head(raw, 2)

# note the raw data is now just a regular R table
class(raw)
```

The `get_code` function is called to check that the processing code is as expected (and for reproducibility).

```{r}
get_code(delayed_data)
```

See the section on [pre-processing Delayed Data](preprocessing-delayed-data.html) to specify additional code instructions to transform your delayed data which will also be added to the output of `get_code`.

#### Aside: Piping functions

The examples above covered some basic piping, but there is a natural sequence to the loading and inspection of a delayed data object. For this reason, the `magrittr` pipe `%>%` works well for many pre-processing tasks.

```{r}
library(teal.data)
library(magrittr)

adsl_cf <- callable_function(function() example_cdisc_data("ADSL"))
cdisc_dataset_connector(
  dataname = "ADSL",
  pull_callable = adsl_cf,
  keys = get_cdisc_keys("ADSL")
) %>%
  mutate_dataset("ADSL$TRTDUR <- round(as.numeric(ADSL$TRTEDTM - ADSL$TRTSDTM), 1)") %>%
  load_dataset() %>%
  get_raw_data() %>%
  head(n = 2)
```

Since these functions modify (operate on) the objects that are given to them, there is no need to assign the result.

For an introduction to pipes, refer to the documentation for `%>%` or other resources on [pipes](https://r4ds.had.co.nz/pipes.html).
